Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-02 response.tex

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

\response{We have included Ji et al.'s work as another alternative method, see Sections 3.2 and 3.3.}

\comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g. ShrinkBayes) on the actual maize dataset in Section 4.  Why is that missing? The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}

\response{We have added a comparison of the computing time required for each method in Section 4 of the revised manuscript. We have also added a comparison of edgeR estimates to eBayes (Laplace) in Figure 3 which show shrinkage for half-parental differences and hybrid effects.}

\response{Added in Section 4. The bottom line is edgeR takes seconds, Ji et al. takes minutes, the other methods take hours.}

@@

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\response{ShrinkBayes failed for us when too many zeros were present. We do not believe the comparison of other methods would change if we did not remove genes with low counts.  In practice, it is common to remove low count genes from RNA-seq datasets prior to analysis because these genes contain little information about differential expression and lead to computational difficulties for some methods. With that being said, we did not remove any genes from the real data analysis.}

\comment{Page 7, lines 17-22.  Can you provide us any information on how well the hyperparameters were estimated based on your empirical Bayes strategy?}

\response{We have assessed the hyperparameter estimates in Section 3.4 and Table 1 where the estimates seem reasonable and improve with increased number of replicates per variety.}

@@

\comment{Page 9, lines 23-25.  In tandem with another reviewer’s comments, the real question is why?  Is it because of the suboptimality of the test that you contrived for ShrinkBayesA?}

\response{In Section 3.3, we highlighted the differences between ShrinkBayes and the eBayes approaches, but do not try to specifically state which difference is the cause.}

\comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}

\response{No, although we do mean ridge rather than peak, i.e. there are many genes with effect size of zero and probabilities below 0.5, and have updated the manuscript accordingly.}

@@

\response{Fixed.}

Reviewer \#1

\comment{The authors develop a hierarchical negative binomial model for drawing inferences from a composite null hypothesis, that tests gene expression heterosis. Simulations are performed to describe the performance of this new method along with a data analysis demonstrating an application to real data. This manuscript represents a well-conceived and executed methodological development in the statistical analysis of RNA-Seq gene expression data. However, the performance of the new method is underwhelming compared to existing methods, tempering my enthusiasm for the likely impact of this paper.}

\response{We assume that you mean the improvement compared to existing methods is underwhelming. To more accurately address the improvement, we calculated percentage improvement in AUC which ranges from 20\% to 100\% and discuss these in Section 3.3.}

@@

\comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al. (2014). Although developed for continuous microarray data, the authors could apply a simple transformation to the count data in order to make them continuous. I would be curious to know whether the eBayes method out-performs the method of Ji et al. (2014) on the transformed data.}

\response{We have assessed the approach of Ji et al. on transformed data, see Sections 3.2 and 3.3. The bottom line is that it doesn't perform as well as the methods based on our hierarchical negative-binomial model.}

\comment{(2) The authors should provide more details on the computational price of the eBayes method? How long (in comparison to edgeR, baySeq, ShrinkBayes) did the eBayes method take to run? Are these methods comparable in terms of computational cost?}

\response{Added in Section 4. The bottom line is edgeR takes seconds, Ji et al. takes minutes, the other methods take hours.}

### 2015-09-02 ebayes_rnaseq_heterosis.tex
hyperparameter estimates

>>>>

\subsection{Hyperparameter estimation}

As described in Section \ref{s:simulation}, we simulated data only from the negative-binomial portion of the model, but not from the hierarchical distributions. Thus, there are no true values for the hyperparameters to compare against. But, we can compute means and standard deviations of the gene-specific parameters that were used in the simulation study and compare these to the estimated hyperparameter values. 
\begin{table}
\caption{Empirical values calculated from the gene-specific parameters used in the simulation and mean of estimated values across the 10 simulations within replicate number. Maximum standard deviation in the estimates across the 10 simulations was 0.004.}
\label{t:hyperparameter}
\begin{tabular}{l|r|rrr|}
& \multicolumn{1}{c|}{Empirical} & \multicolumn{3}{c}{\# of replicates}\\
Parameter & value & 4 & 8 & 16 \\
\hline
% with (SD)
% $\eta_\phi$ & 4.675 & 4.681 (0.004) & 4.741 (0.002) & 4.817 (0.003) \\
% $\eta_\alpha$ & -0.017 & -0.023 (0.002) & -0.018 (0.003) & -0.014 (0.002)  \\
% $\eta_\delta$ & -0.046 & -0.046 (0.002) & -0.031 (0.002) & -0.022 (0.003)  \\
% $\eta_\phi$ & -2.713 & -2.290 (0.004) &  -2.335 (0.003) & -2.387 (0.002)\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821 (0.002) & 1.779 (0.002) & 1.726 (0.003)\\
% $\sigma_\alpha$ & 0.364 & 0.459 (0.002) & 0.399 (0.002) & 0.357 (0.002) \\
% $\sigma_\delta$ & 0.145 & 0.389 (0.002) & 0.317 (0.002) & 0.268 (0.001) \\
% $\sigma_\psi$ & 0.737 & 0.401 (0.001) & 0.493 (0.001) & 0.542 (0.001) \\
% 3 decimal places
% $\eta_\phi$ & 4.675 & 4.681 & 4.741  & 4.817  \\
% $\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
% $\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
% $\eta_\phi$ & -2.713 & -2.290  &  -2.335  & -2.387\\
% \hline
% $\sigma_\phi$ & 1.815 & 1.821  & 1.779  & 1.726 \\
% $\sigma_\alpha$ & 0.364 & 0.459  & 0.399  & 0.357  \\
% $\sigma_\delta$ & 0.145 & 0.389  & 0.317  & 0.268  \\
% $\sigma_\psi$ & 0.737 & 0.401 & 0.493 & 0.542 \\
$\eta_\phi$ & 4.7\phantom{00} & 4.7\phantom{00} & 4.7\phantom{00}  & 4.8\phantom{00}  \\
$\eta_\alpha$ & -0.017 & -0.023  & -0.018 & -0.014  \\
$\eta_\delta$ & -0.046 & -0.046 & -0.031  & -0.022   \\
$\eta_\phi$ & -2.7\phantom{00} & -2.3\phantom{00}  &  -2.3\phantom{00}  & -2.4\phantom{00} \\
\hline
$\sigma_\phi$ & 1.8\phantom{0} & 1.8\phantom{0}  & 1.8\phantom{0}  & 1.7\phantom{0} \\
$\sigma_\alpha$ & 0.36 & 0.46  & 0.40  & 0.36  \\
$\sigma_\delta$ & 0.15 & 0.39  & 0.32  & 0.27  \\
$\sigma_\psi$ & 0.74 & 0.40 & 0.49 & 0.54 \\
\hline
\end{tabular}
\end{table}
Table \ref{t:hyperparameter} provides the computed means and standard deviations from the gene-specific parameter values, called empirical values, and compares these values against the mean of estimated hyperparameters across the 10 simulations within replicate number. For the mean parameters, there is reasonable agreement amongst the different numbers of replicates with a somewhat high estimate for the mean of the overdispersion parameter $(\eta_\phi)$ and, perhaps, a trend for the mean of the hybrid effect ($\eta_\delta$), although we would not have expected this. For the standard deviation parameters, $\sigma_\phi$ is consistent across the replicate numbers, $\sigma_\alpha$ and $\sigma_\delta$ mean estimates are decreasing toward the empirical values as the number of replicates increases, and $\sigma_\phi$ is increasing toward the empirical value as the number of replicates increases. %We expect this pattern for $\sigma_\alpha$ and $\sigma_\delta$ since these are based on moment matching of independent gene-specific parameter estimates. For low numbers of replicates, these independent estimates will have relatively large sampling variability that will cause a bias toward larger values for the moment-matched standard deviation estimates.
